Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-01 paper.Rnw
Responded to all of mallar's comments

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
nonlinear image registration. The resultant segmentations are then propagated
to each target image and fused using a label fusion method.  To explore the
effect of atlases and templates library size, registation and label
fusion method, we conduct a 10-fold Monte Carlo cross-validation of \mb{} of whole
hippocampal segmentation on 60 ADNI subjects, manually segmented using the
\prsnr{}-protocol, over a range of parameter settings. Using the best settings
found, we then conduct a leave-one-out cross-validation (LOOCV) of hippocampal
subfield segmentation using five high-resolution manually segmented atlases.
Two final experiments assess \mb{} when applied to first episode psychosis and
Alzheimer's disease populations, and \mb{} segmentations are compared with
existing automated methods (\fsl{}, \freesurfer{}, \maper{}, and \snt) and
biases are explored.\\
\textbf{Results:} \textbf{[TODO: update with FEP-XVAL and subfield XVAL results.]} 
Using 9 atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of \Sexpr{mb.kappa.best} 
(to \psegs{}) over 10-folds of Monte Carlo 
significantly lower variability in DSC than \multiatlas{} segmentation.
In a LOOCV, \mb{} accurately reproduces the subfields... .

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                 TODO 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\parbox{7in}{
\subsection{TODO:}
\begin{enumerate}
\itemsep-0.5em
\item check tense: we are is have has use/used find/found
\item remove 'accuracy' unless specifically talking relative to ground truth (and
    even then, qualify)
\item change overlap/similarity --> agreement? choose one. 
\item check that all chart labels are sensible 
\item check that Experiment 5 matches Experiment 1
\item check that all Experiment references resolve correctly
\item fix figure 2 schematic language for the experiments (not 2a 2b). 
\item ADNI-1 or ADNI1: standardize
\item check that experiment references are formatted as Experiment 2 (Section 3.2)
\item subfield or subregion -- choose one
\end{enumerate}}

@@

Of particular relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas{} segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic \multiatlas{} strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas{} techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against semi-automated segmentations.

@@

\begin{itemize}
  \item \expadnixval{} investigates the accuracy of \mb{} whole hippocampus
  segmentation of aging and Alzhiemer's diseased subjects over a wide range of
  parameter settings using a Monte Carlo cross-validation design. The results of
  this experiment enable us to choose the parameter settings offering the best
  performance for use in subsequent experiments.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<tab:ADNI-xval-demographics, echo=F, dependson='ADNI1-xval-load', results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlases$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Evaluation method}  
The Dice similarity coefficient (DSC), also known as Dice's Kappa, assesses the
agreement between two segmentations. It is one of the most widely used measures
of segmentation accuracy, and we use it as the basis of comparison in this
experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}).  A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion).  The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation accuracy, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation.  Given
these findings, in the remainder of our experiments only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

@@

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (\expadnixvalsnt{}, Supplementary Materials) where a different
hippocampal definition is used (\snt labels available with the ADNI datasets).
This strongly suggests that these results are independant of the segmentation
protocol used and are, instead, features of the \mb{} algorithm.

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI1-xval-k-diff}
    \caption{Increase in similarity score over \multiatlas} 
    \label{fig:ADNI1-xval-k-diff}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI1-xval-variability}
    \caption{Difference in variability} 
    \label{fig:ADNI1-xval-variability}
  \end{subfigure}
  \caption{\textbf{Whole hippocampus segmentation cross-validation on
  Alzheimer's diseased subjects.}
(\ref{fig:ADNI1-xval-k-diff}) shows increase in mean Dice's Similarity Coefficient of \mb{} over
  \multiatlas{} segmentations vs. atlas and template library size when using the
  \ants{} registration method, and majority-vote label fusion over 10 folds of
  Monte Carlo cross validation.  Segmentation similarity is computed against
  \psegs{}.  Error bars indicate standard error.
(\ref{fig:ADNI1-xval-variability}) compares the variability in Dice scores of \mb{} vs.
  \multiatlas{} segmentations in the validation.  Variance of segmentation
  accuracy between \mb{} and \multiatlas{} segmentation is computed for each
  subject across over 10 folds of Monte Carlo cross validation.  Shown on
  the y-axis is the p-value resulting from a t-test comparing the distribution of
  variances at each parameter setting (atlas/template library size).  Only points
  where \mb{} mean variability is lower than \multiatlas{} are shown.  Dashed
  lines indicate a p-value of 0.05 and 0.01.}
\end{figure}

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of Psychosis patients                         %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expfepxval{}: \expfepxvaldesc{}}
\label{exp:fep-xval}
\textbf{[TODO: tie with exp 1]}
To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases
\citep{Winterburn2013} to derive whole hippocampal segmentations of a dataset of
patients having had a single episode of psychosis. The resulting segmentations
are assessed for quality by comparison with expert manual segmentations.

\subsubsection{\expfepxval{}: Materials and Methods}
\paragraph{First Episode Psychosis Dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 14 to 35 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

Manual segmentation of each subject whole hippocampus is produced following a
validated segmentation protocol \citep{Pruessner2000}.

@@

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} and two other popular segmentation packages, \fsl{} and
\freesurfer{}. Firstly, \mb{} is used with the five Winterburn atlas images and labels as atlases, and a
randomly selected subset of 19 psychosis dataset images as templates. Secondly,
\mb{} is run using the same template images, but instead as atlases we use five
additional five first episode psychosis subjects and corresponding manual
segmentations (not included above). Finally, \fsl{} and \freesurfer{} are run
with their default settings to segment the entire dataset. The \fsl
\verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer was run using \verb+recon-all+. \textbf{

@@

The Pruessner segmentation protocol differs slightly from the Winterburn
protocol those used by \freesurfer and \fsl in the neuroanatomical features that
are delineated (see \citet{Winterburn2013}, and Table
\label{tab:subfield-segmentation-protocols} in the Discussion below).  This
variation in protocol poses a problem if an overlap measure is used for
evaluation: since different protocols will necessarily produce segmentations
that do not perfectly overlap, the degree of overlap cannot be solely used to
compare segmentation methods using different protocols. In place of an overlap
metric, we assess the degree of (Pearson) correlation in average bilateral
hippocampal volume produced by each method. Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 2, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of psychosis patients                         %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\expfepxval{}: Results}
\textbf{[TODO: move up results to here.]}

% In this experiment \mb{} is applied to a dataset of images of first episode
% psychosis patients, using the Winterburn atlases and a template library of
% 21 subject images selected at random.  Expert manual whole hippocampal
% segmentations are used as a gold standard.

@@

<<fig:FEP-Bland-Altman,dependson='fig:FEP-vols',fig.width=7,fig.height=3,fig.show='hide'>>=
fep.volumes$diff = fep.volumes$manual - fep.volumes$volume
fep.volumes$mean = (fep.volumes$manual + fep.volumes$volume)/2
limits = ddply(fep.volumes, c("Method"), function (df) { 
  data.frame(
    y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T) 
)})

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expadni{}: \expadnidesc{}}
\label{exp:adni}
To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated
whole hippocampal segmentations (\snt{}) provided by ADNI.

\subsubsection{\expadni{}: Materials and Methods}
\paragraph{\adnidataset{} dataset} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

\paragraph{Experiment details} 
\mb{} was configured with an atlas library composed of the five Winterburn T1
atlases (described in \expsubfield{}, section \ref{exp:winterburn-xval}), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in \expadnixval{} (Section \ref{exp:adni-xval}).  The \fsl{} segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset{} dataset were segmented by both methods.

One author (MP) performed visual quality inspection for \mb{} and \fsl{} 
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (\snt{}, \maper{}, \freesurfer{}, \fsl{}, and \mb{}) and also
passing quality control inspection were included in the analysis.

\paragraph{Evaluation method}
As in previous experiments, the \snt{} and Winterburn hippocampal segmentation protocols
differ in the neuroanatomical features delineated, and so we assess \mb{} by
the degree of (Pearson) correlation of average hippocampal volume across
subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl{}, \freesurfer{},
and \maper{}, and \snt{} semi-automated segmentations. Additionally, we evaluate
the volume-related fixed and proportional biases in all segmentation methods
using Bland-Altman plots \citep{Bland1986}.

@@

@@

Based on the results from \expadnixval{} (Section \ref{exp:adni-xval}), in this
experiment \mb{} was configured with a template library of 21 randomly chosen
subject images (7 from each disease class) and majority vote label fusion.  The
entire \adnidataset{} dataset was segmented by \mb{}, and the resulting volumes
compared with those obtained by expert semi-automated segmentation (\snt{}), and
three other automated segmentation techniques: \maper{}, \freesurfer{}, and
\fsl{}. Table \ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the
total count of segmentations available, including a count of those which have
failed a quality control inspection.  A total of
\Sexpr{length(means.complete$RID)} images are included in the following
analysis, having met quality criteria and having segmentations from every
method.

@@

\FloatBarrier

>>>>

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches. These approaches often rely on
a large number of such manually segmented atlases (e.g. 30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar aggreement to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
nonlinear image registration. The resultant segmentations are then propagated
to each target image and fused using a label fusion method. To explore the
effect of atlases and templates library size, registation and label
fusion method, we conduct a 10-fold Monte Carlo cross-validation of \mb{} of whole
hippocampal segmentation on 60 ADNI subjects, manually segmented using the
\prsnr{}-protocol, over a range of parameter settings. Using the best settings
found, we then conduct a leave-one-out cross-validation (LOOCV) of hippocampal
subfield segmentation using five high-resolution manually segmented atlases.
Two final experiments assess \mb{} when applied to first episode psychosis and
Alzheimer's disease populations, and \mb{} segmentations are compared with
existing automated methods (\fsl{}, \freesurfer{}, \maper{}, and \snt) and
biases are explored.\\
\textbf{Results:} \textbf{[TODO: update with FEP-XVAL and subfield XVAL results.]} 
Using 9 atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of \Sexpr{mb.kappa.best} 
(to \psegs{}) over 10-folds of Monte Carlo 
significantly lower variability in DSC than \multiatlas{} segmentation.
In a LOOCV, \mb{} accurately reproduces the subfields... .

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                 TODO 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\parbox{7in}{
\subsection{TODO:}
\begin{enumerate}
\itemsep-0.5em
\item check tense: we are is have has use/used find/found
\item remove 'accuracy' unless specifically talking relative to ground truth (and
    even then, qualify)
\item change overlap/similarity --> agreement? choose one. 
\item fix figure 2 schematic language for the experiments (not 2a 2b). 
\item check that experiment references are formatted as Experiment 2 (Section 3.2)
\end{enumerate}}

@@

Of particular relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas{} segmentation with a limited number of input atlases. The
LEAP algorithm is a clever modification to the basic \multiatlas{} strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas{} techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library. In their validation, LEAP was used to segment the whole
hippocampus in the ADNI1 baseline dataset, achieving a mean Dice score of 0.85
against semi-automated segmentations.

@@

\begin{itemize}
  \item \expadnixval{} investigates \mb{} whole hippocampus
  segmentation of aging and Alzhiemer's diseased subjects over a wide range of
  parameter settings using a Monte Carlo cross-validation design. The results of
  this experiment enable us to choose the parameter settings offering the best
  performance for use in subsequent experiments.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<tab:ADNI-xval-demographics, echo=F, dependson='ADNI1-xval-load', results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlases$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease. 
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes. 
  ADAS   - Alzheimer's Disease Assessment Scale. 
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Evaluation method}  
The Dice similarity coefficient (DSC), also known as Dice's Kappa, assesses the
agreement between two segmentations. It is one of the most widely used measures
of segmentation agreement, and we use it as the basis of comparison in this
experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion). The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation agreement, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation. Given
these findings, in the remainder of our experiments only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

@@

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (\expadnixvalsnt{}, Supplementary Materials) where a different
hippocampal definition is used (\snt{} labels available with the ADNI datasets).
This strongly suggests that these results are independant of the segmentation
protocol used and are, instead, features of the \mb{} algorithm.

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI1-xval-k-diff}
    \caption{Increase in similarity score over \multiatlas} 
    \label{fig:ADNI1-xval-k-diff}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI1-xval-variability}
    \caption{Difference in variability} 
    \label{fig:ADNI1-xval-variability}
  \end{subfigure}
  \caption{\textbf{Whole hippocampus segmentation cross-validation on
  Alzheimer's diseased subjects.}
(\ref{fig:ADNI1-xval-k-diff}) shows increase in mean Dice's Similarity Coefficient of \mb{} over
  \multiatlas{} segmentations vs. atlas and template library size when using the
  \ants{} registration method, and majority-vote label fusion over 10 folds of
  Monte Carlo cross validation. Segmentation similarity is computed against
  \psegs{}. Error bars indicate standard error.
(\ref{fig:ADNI1-xval-variability}) compares the variability in Dice scores of \mb{} vs.
  \multiatlas{} segmentations in the validation. Variance of segmentation
  agreement between \mb{} and \multiatlas{} segmentation is computed for each
  subject across over 10 folds of Monte Carlo cross validation. Shown on
  the y-axis is the p-value resulting from a t-test comparing the distribution of
  variances at each parameter setting (atlas/template library size). Only points
  where \mb{} mean variability is lower than \multiatlas{} are shown. Dashed
  lines indicate a p-value of 0.05 and 0.01.}
\end{figure}

%
%-------------------------------------------------------------------------------
%

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of Psychosis patients                         %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expfepxval{}: \expfepxvaldesc{}}
\label{exp:fep-xval}
To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we replicate the cross-validation done in
\expadnixval{} a dataset of patients having had a single episode of psychosis. We
also compare \mb{} segmentations with those of two well-known automated
segmentation methods, \fsl{} and \freesurfer{}.

\subsubsection{\expfepxval{}: Materials and Methods}
\paragraph{First Episode Psychosis (FEP) Dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 14 to 35 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis. For complete program details see
\citet{Malla2003}.

@@

Export whole hippocampal manual segmentation of each subject is produced
following a validated segmentation protocol \citep{Pruessner2000}.

@@

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} and two other popular segmentation packages, \fsl{} and
\freesurfer{}. Firstly, \mb{} is used with the five Winterburn atlas images and
labels as atlases, and a randomly selected subset of 19 psychosis dataset images
as templates. Secondly, \mb{} is run using the same template images, but instead
as atlases we use five additional five first episode psychosis subjects and
corresponding manual segmentations (not included above). Finally, \fsl{} and
\freesurfer{} are run with their default settings to segment the entire dataset.
The \fsl{} \verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer{} was run using \verb+recon-all -all+.

@@

The Pruessner segmentation protocol differs slightly from the Winterburn
protocol those used by \freesurfer{} and \fsl{} in the neuroanatomical features
that are delineated (see \citet{Winterburn2013}, and Table
\ref{tab:subfield-segmentation-protocols} in the Discussion below). This
variation in protocol poses a problem if an overlap measure is used for
evaluation: since different protocols will necessarily produce segmentations
that do not perfectly overlap, the degree of overlap cannot be solely used to
compare segmentation methods using different protocols. In place of an overlap
metric, we assess the degree of (Pearson) correlation in average bilateral
hippocampal volume produced by each method. Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

%
%-------------------------------------------------------------------------------
%

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 2, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of psychosis patients                         %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\expfepxval{}: Results}

As in \expadnixval, we find that similarity score increases with a greater
number of atlases or templates but quickly plateaus (Figure
\label{fig:FEP-xval}). A maximum similarity score of \Sexpr{snt.mb.kappa.best}
is found when using 9 atlases, 19 templates, ANTS registration, and majority
vote label fusion.

We found a close relationship in average hippocampal volume between the manual
label volumes and \mb{} when using the Winterburn atlases, or manually segmented
FEP subjects as atlases (Figure \label{fig:FEP-vols}). Both sets of volumes are
correlated with Pearson $r > 0.88$. \freesurfer{} and \fsl{} volumes are both
correlated with manual volumes at Pearson $r > 0.7$.

As \citet{Bland1986} noted, high correlation amongst measures of the same
quantity does not necessarily imply agreement (as correlation can be driven by a
large range in true values, for instance). Figure \label{fig:FEP-Bland-Altman}
shows Bland-Altman plots illustrating the level of agreement of each method with
manual volumes. All methods show an obvious proportional bias: \freesurfer{}
and \fsl{} markedly underestimate smaller hippocampi and over-estimate large
hippocampi (the limits of agreement are between $-2482^3mm$ and $-784^3mm$, and
between $-1653^3mm$ and $79^3mm$, respectively), whereas both \mb{} methods show
a much less exaggerated, but conservative bias (limits of agreement between
$-67^3mm$ and $766^3mm$ when using FEP atlases, and between $-333^3mm$ and
$504^3mm$ when using Winterburn atlases). On average, \freesurfer{} and \fsl{}
overestimate hippocampal volume by about $1600 mm^3$ and $800 mm^3$,
respectively. In contrast, on average \mb{} underestimates volumes by about $300
mm^3$ when using FEP atlases, and by about $80mm^3$ when using Winterburn
atlases.

@@

<<fig:FEP-Bland-Altman,dependson='fig:FEP-vols',fig.width=10,fig.height=3,fig.show='hide'>>=
fep.volumes$diff = fep.volumes$manual - fep.volumes$volume
fep.volumes$mean = (fep.volumes$manual + fep.volumes$volume)/2
limits = ddply(fep.volumes, c("Method"), function (df) { 
  data.frame(
    y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T) 
)})

@@

%
%-------------------------------------------------------------------------------
%

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expadni{}: \expadnidesc{}}
\label{exp:adni}
To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to a large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated whole
hippocampal segmentations (\snt{}) provided by ADNI.

%%%%%%%%%%
\subsubsection{\expadni{}: Materials and Methods}

\paragraph{\adnidataset{} dataset}  %%%%%%%%%%%
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper{}, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

\paragraph{Winterburn Atlases}  %%%%%%%%%%
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $0.3mm$-isotropic T1-weighted MR images, described
more fully in \expfepxval, section \ref{exp:fep-xval}.

\paragraph{Experiment details} %%%%%%%%%%
\mb{} was configured with an atlas library composed of the five Winterburn and
segmentations. A template library of 19 images were randomly selected from the target
dataset of ADNI subjects, and \ants{} registration and majority vote label fusion
were used as these were found to perform favourably in earlier experiments.

\fsl{} segmentation was performed using the \verb+run_first_all+ script
according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}. All images in
the \adnidataset{} dataset were segmented by both methods.

One author (MP) performed visual quality inspection for \mb{} and \fsl{}
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass). Only images meeting the conditions of having
segmentations from all methods (\snt{}, \maper{}, \freesurfer{}, \fsl{}, and
\mb{}) and also passing similary quality control inspection were included in the
analysis.

\paragraph{Evaluation method} %%%%%%%%%%
As in previous experiments, the Winterburn hippocampal segmentation protocol
differs in the delineated neuroanatomical features (\citet{Winterburn2013}, and
Table \ref{tab:subfield-segmentation-protocols}, Discussion) and so we assess
\mb{} by the degree of (Pearson) correlation of average hippocampal volume
across subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl{}, \freesurfer{},
and \maper{}, and \snt{} semi-automated segmentations. Additionally, we evaluate
the volume-related fixed and proportional biases in all segmentation methods
using Bland-Altman plots \citep{Bland1986}.

@@

%
%-------------------------------------------------------------------------------
%

@@

@@

%
%-------------------------------------------------------------------------------
%
